Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}

### 2016-02-09 jss/article.tex
[skip ci] update table

%% almost as usual
\author{Paul Deveau\\Institut Curie\\PSL University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
        Emmanuel Barillot\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And 
        Valentina Boeva\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \AND 
        Andrei Zinovyev\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And
        Eric Bonnet\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech}
\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}

@@

The example shown below is the simplest command that can be done to test a gene
list for over-representation on the six included ACSN maps. With the list of 12
genes mentionned above and a default p-value cutoff of $0.05$, we have a set of
36 maps or modules that are significantly enriched. The results are structured
as a data frame with nine columns displaying the module name, the module size,
the number of genes from the list in the module, the names of the genes that are
present in the module, the size of the reference universe, the number of genes
from the list that are present in the universe, the raw p-value, the p-value
corrected for multiple testing and the type of test performed. The module field
in the results data frame indicate the map name and the module name separated by
a column character. If a complete map is significantly enriched or depleted,
then only the map name is shown, without any module or column character. For
instance, the first line of the results object below concern the E2F1 module of
the Apoptosis map.

\begin{verbatim}
> library(ACSNMineR)
> results <- enrichment(genes_test)
> dim(results)
[1] 36  9
> results[1,]
             module module_size nb_genes_in_module    genes_in_module 
V17 APOPTOSIS_ENTRY          10                  4 ATM ATR CHEK2 E2F1          
    universe_size nb_genes_in_universe      p.value	p.value.corrected    test
V17      2241                   12 3.491372e-07	6.400849e-06 greater
\end{verbatim}

@@

The dataset corresponding to this study is available as a Bioconductor package.
The code shown below is creating a list of differentially expressed genes
between ER positive and ER negative samples, and calculates the enrichment in
ACSN maps from this list of genes. As seen in Table \ref{tab:table_mainz}, there
are four maps (Apoptosis, DNA repair, ACSN master and Survival) and seven
modules enriched. Most of the modules are in fact linked to different phases of
the cell cycle.

@@

\begin{tabular}{l|c|c|c|c}
\hline
Module & Mod. size & Nb genes & pval & Cor. pval\\
\hline
Apoptosis:HEDGEHOG & 31 & 21 & 0.0016 & 0.042 \\
Apoptosis:AKT\_MTOR &  79 & 47 & 0.00037 & 0.038 \\
DNA\_repair & 335 & 163 & 0.0012 & 0.038 \\
DNA\_repair:G1\_CC\_PHASE &  25 & 18 & 0.0011 & 0.038 \\
ACSN\_master:AKT\_MTOR & 92 & 53 & 0.0005 & 0.038 \\
ACSN\_master:G1\_CC\_PHASE & 25 & 18 & 0.0011 & 0.038 \\
Survival:S\_PHASE\_CHECKPOINT & 8 & 8 & 0.0006 & 0.038 \\
\hline
\end{tabular}
\end{table}

>>>>

%% almost as usual
\author{Paul Deveau\\Institut Curie\\PSL University\\Univ. Paris-Sud\\INSERM U830\\INSERM U900\\Mines-ParisTech \And 
        Emmanuel Barillot\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And 
        Valentina Boeva\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech\\Institut Cochin\\INSERM U1016\\CNRS UMR 8104\\Paris Descartes University UMR-S1016 \AND 
        Andrei Zinovyev\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech \And
        Eric Bonnet\\Institut Curie\\PSL University\\INSERM U900\\Mines-ParisTech}
\title{Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with the R packages \pkg{ACSNMineR} and \pkg{RNaviCell}}

@@

The example shown below is the simplest command that can be done to test a gene
list for over-representation on the six included ACSN maps. With the list of 12
genes mentionned above and a default p-value cutoff of $0.05$, we have a set of
36 maps or modules that are significantly enriched. The results are structured
as a data frame with nine columns displaying the module name, the module size,
the number of genes from the list in the module, the names of the genes that are
present in the module, the size of the reference universe, the number of genes
from the list that are present in the universe, the raw p-value, the p-value
corrected for multiple testing and the type of test performed. The module field
in the results data frame indicate the map name and the module name separated by
a column character. If a complete map is significantly enriched or depleted,
then only the map name is shown, without any module or column character. For
instance, the third line of the results object below concern the E2F1 module of
the CellCycle map.

\begin{verbatim}
> library(ACSNMineR)
> results <- enrichment(genes_test)
> dim(results)
[1] 8  9
> results[3,]
            module module_size nb_genes_in_module
V161 CellCycle:E2F1          19                 12
                                                             genes_in_module
V161 ATM ATR CHEK2 CREBBP TFDP1 E2F1 EP300 HDAC1 KAT2B GTF2H1 GTF2H2 GTF2H2B
     universe_size nb_genes_in_universe      p.value p.value.corrected    test
V161          2237                   12 3.735018e-21      2.353061e-19 greater
\end{verbatim}

@@

The dataset corresponding to this study is available as a Bioconductor package.
The code shown below is creating a list of differentially expressed genes
between ER positive and ER negative samples, and calculates the enrichment in
ACSN maps from this list of genes. As seen in Table \ref{tab:table_mainz},
there is one map (Cell Cycle) and six modules (belonging to the Cell Cycle, DNA
repair and EMT motility maps) enriched.

@@

\begin{tabular}{l|c|c|c|c}
\hline
Module & Mod. size & Nb genes & pval & Cor. pval\\
\hline
CellCycle & 242 & 20 & 1.70e-03 & 0.021 \\
CellCycle:E2F1\_TARGETS & 130 & 14 & 6.26e-04 & 0.009 \\
CellCycle:E2F2\_TARGETS & 35 & 8 & 4.70e-05 & 0.009 \\
CellCycle:E2F3\_TARGETS & 51 & 11 & 3.82e-06 & 0.0002 \\
CellCycle:E2F4\_TARGETS & 100 & 15 & 9.99e-06 & 0.0003 \\
DNA\_repair:SPINDLE\_CHECKPOINT & 28 & 5 & 3.82e-03 & 0.04 \\
EMT\_motility:ECM & 147 & 13 & 5.09e-03 & 0.045 \\
\hline
\end{tabular}
\end{table}
